---
title: "Gene_length_effect"
author: "Lauren Blake"
date: "February 21, 2017"
output: html_document
---

The goal of this script is to check the relationship between the length of a gene and its propensity to be called DE. 

```{r}
# Load libraries

library("gplots")
library("ggplot2")
source("~/Desktop/Endoderm_TC/ashlar-trial/analysis/chunk-options.R")
library("colorfulVennPlot")
library("VennDiagram")

# Load data

length_diff <- read.delim("~/Dropbox/Endoderm TC/length_diff.txt")
```

## Basic plot of length difference by DE status 
```{r}
# For each day

p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot()

p


# Each day plotted individually

p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) + ylab("Human length -  chimp length (bp)") + xlab("DE (5% FDR)")

p

q <- p + facet_wrap( ~ day) 
q

# Zoom in

p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) 

q <- p + facet_wrap( ~ day) + ylim(c(-50,50))
q

```

## Genes with difference > 10bp between species

```{r}

length_diff_10 <- length_diff[which(length_diff$abld > 10) , ]

# Each day plotted individually

p <- ggplot(length_diff_10, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) + ylab("Human length -  chimp length (bp)") + xlab("DE (5% FDR)")

q <- p + facet_wrap( ~ day) +geom_jitter(position=position_jitter(width=.1, height=0)) + ylim(c(-200,200))
q


```

## Genes with difference > 10bp between species further divided into direction of effect size

```{r}
# Separate by direction

length_diff_10$logFC[length_diff_10$logFC <= 0] <- -1
length_diff_10$logFC[length_diff_10$logFC > 0] <- 1

length_diff_10$inter <- interaction(length_diff_10$day, length_diff_10$logFC)

# Each day plotted individually

p <- ggplot(length_diff_10, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=inter)) + ylab("Human length -  chimp length (bp)") + xlab("DE at 5% FDR (neg. logFC / pos. logFC)")

q <- p + facet_wrap( ~ day) +geom_jitter(position=position_jitter(width=0.8, height=1)) + ylim(c(-200,200))
q
```

## RPKM

```{r}

# Get orth exon lengths

ortho_exon_lengths <- read.delim("~/Dropbox/Endoderm TC/ortho_exon_lengths.txt")

totalNumReads <- as.data.frame(t(colSums(counts_genes_in_cutoff, na.rm = FALSE, dims = 1) ))

# Calculate per species RPKM

humans <- c(1:7, 16:23, 32:39, 48:55)
chimps <- c(8:15, 24:31, 40:47, 56:63)
   
  # humans         
#RPKM_humans_trans <- as.data.frame(counts_genes_in_cutoff[1:10304, humans] * 1000000 / (ortho_exon_lengths$hutotal /1000))

#humans_reads <-  totalNumReads[,humans]

#RPKM_humans[,1:31] <-   RPKM_humans_trans[,1:31] / humans_reads

# Make RPKM into a row

RPKM_humans <- rpkm(counts_genes_in_cutoff, gene.length=ortho_exon_lengths$hutotal/1000, normalized.lib.sizes=TRUE, log=TRUE) 

RPKM_chimps <- rpkm(counts_genes_in_cutoff, gene.length=ortho_exon_lengths$chtotal/1000, normalized.lib.sizes=TRUE, log=TRUE) 

# Take human samples from the human RPKM and chimp samples from the chimp RPKM data frames

RPKM_all <- cbind(RPKM_humans[,1:7], RPKM_chimps[,8:15], RPKM_humans[,16:23], RPKM_chimps[,24:31], RPKM_humans[,32:39], RPKM_chimps[,40:47], RPKM_humans[,48:55], RPKM_chimps[,56:63])

# Calculate the Pearson's correlation for each sample

Cor_values = matrix(data = NA, nrow = 63, ncol = 1, dimnames = list(c("human 0", "human 0", "human 0", "human 0", "human 0", "human 0", "human 0", "chimp 0", "chimp 0", "chimp 0", "chimp 0", "chimp 0", "chimp 0", "chimp 0", "chimp 0", "human 1", "human 1", "human 1", "human 1", "human 1", "human 1", "human 1", "human 1", "chimp 1", "chimp 1", "chimp 1", "chimp 1", "chimp 1", "chimp 1", "chimp 1", "chimp 1",  "human 2", "human 2", "human 2", "human 2", "human 2", "human 2",  "human 2", "human 2",  "chimp 2", "chimp 2", "chimp 2", "chimp 2", "chimp 2", "chimp 2", "chimp 2", "chimp 2", "human 3", "human 3", "human 3", "human 3", "human 3", "human 3", "human 3", "human 3", "chimp 3",  "chimp 3",  "chimp 3",  "chimp 3", "chimp 3",  "chimp 3",  "chimp 3",  "chimp 3"), c("Pearson's correlation")))

for (i in 1:63){
  Cor_values[i,1] <- cor(RPKM_all[,i], cpm_cyclicloess[,i])
}

summary(Cor_values)
```

## Check RPKM 

```{r}
kb_gene_length_human <- ortho_exon_lengths$hutotal / 1000

kb_gene_length_chimp <- ortho_exon_lengths$chtotal /1000

gene_length_all <- cbind(kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human,
                        kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp,
                         kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, 
                          kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp,
                              kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human,
                         kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp,
                         kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human, kb_gene_length_human,
                         kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp, kb_gene_length_chimp)
```




## Here, we want to identify DE genes across species and timepoints with an individual random effect (Set 1)

```{r}

# Make the dge object

# Load the data 

gene_counts_cutoff_norm_data <- read.delim("~/Desktop/Endoderm_TC/ashlar-trial/data/gene_counts_cutoff_norm_data.txt")

Endoderm_mapping_core_1 <- read.csv("~/Desktop/Endoderm_TC/Endoderm_mapping_core_1.csv")

# Make labels with species and day

individual <- After_removal_sample_info$Individual

species <- c("H", "H","H","H","H","H","H", "C", "C","C","C","C","C","C","C","H","H","H","H","H","H","H","H",  "C", "C","C","C","C","C","C","C", "H","H","H","H","H","H","H","H",  "C", "C","C","C","C","C","C","C", "H","H","H","H","H","H","H","H",  "C", "C","C","C","C","C","C","C")

day <- c("0", "0","0","0","0","0","0", "0", "0", "0","0","0","0","0", "0", "1","1","1","1","1","1","1","1", "1","1","1","1","1","1","1","1",  "2", "2","2","2","2","2","2","2","2", "2","2","2","2","2","2","2",  "3", "3","3","3","3","3","3","3",  "3", "3","3","3","3","3","3", "3")

labels <- paste(species, day, sep=" ")

# Take the TMM of the genes that meet the criteria

dge_in_cutoff <- DGEList(counts=as.matrix(gene_counts_cutoff_norm_data), genes=rownames(gene_counts_cutoff_norm_data), group = as.character(t(labels)))
dge_in_cutoff <- calcNormFactors(dge_in_cutoff)



design <- model.matrix(~ species*day )

colnames(design)[1] <- "Intercept"
colnames(design) <- gsub("speciesH", "Human", colnames(design))
colnames(design) <- gsub(":", ".", colnames(design))


# We want a random effect term for individual. As a result, we want to run voom twice. See https://support.bioconductor.org/p/59700/


cpm.voom <- voom(dge_in_cutoff, design, normalize.method="cyclicloess")

corfit.correlation = corfit$consensus.correlation
cpm.voom.corfit <- voom(dge_in_cutoff, design, plot = TRUE, normalize.method="cyclicloess", block=individual, correlation = corfit.correlation )

# Replace expression with RPKM values in the dge object
#cpm.voom.corfit$E <- cpm.voom.corfit$E - log2(gene_length_all)

cpm.voom.corfit$E <- as.data.frame(RPKM_all)
RPKM_vals <- cpm.voom.corfit$E


pca_genes <- prcomp(t(RPKM_vals), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
pc1 <- matrixpca[,1]
pc2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]

pcs <- data.frame(pc1, pc2, pc3, pc4, pc5)

summary <- summary(pca_genes)

#dev.off()

ggplot(data=pcs, aes(x=pc1, y=pc2, color=day, shape=Species, size=2)) + geom_point(aes(colour = as.factor(day))) +  scale_colour_manual(name="Day",  
                      values = c("0"=rgb(239/255, 110/255, 99/255, 1), "1"= rgb(0/255, 180/255, 81/255, 1), "2"=rgb(0/255, 177/255, 219/255, 1),
                                 "3"=rgb(199/255, 124/255, 255/255,1))) + xlab(paste("PC1 (",(summary$importance[2,1]*100),"% of variance)")) + ylab(paste("PC2 (",(summary$importance[2,2]*100),"% of variance)")) + scale_size(guide = 'none')  + theme_bw()  + ggtitle("PCs 1 and 2 from TMM normalized RPKM (63 samples)")
     

# Run lmFit and eBayes in limma

fit <- lmFit(cpm.voom.corfit , design, block=individual, correlation=corfit.correlation)


# In the contrast matrix, we have the species DE at each day

cm2 <- makeContrasts(HvCday0 = Human, HvCday1 = Human + Human.day1, HvCday2 = Human + Human.day2, HvCday3 = Human + Human.day3, Hday01 = day1 + Human.day1, Hday12 = day2 + Human.day2 - day1 - Human.day1, Hday23 = day3 + Human.day3 - day2 - Human.day2, Cday01 = day1, Cday12 = day2 - day1, Cday23 = day3 - day2, Sig_inter_day1 = Human.day1, Sig_inter_day2 = Human.day2 - Human.day1, Sig_inter_day3 = Human.day3 - Human.day2, levels = design)

# Fit the new model
diff_species <- contrasts.fit(fit, cm2)
fit2 <- eBayes(diff_species)

top3 <- list(HvCday0 =topTable(fit2, coef=1, adjust="BH", number=Inf, sort.by="none"), HvCday1 =topTable(fit2, coef=2, adjust="BH", number=Inf, sort.by="none"),  HvCday2 =topTable(fit2, coef=3, adjust="BH", number=Inf, sort.by="none"),  HvCday3 =topTable(fit2, coef=4, adjust="BH", number=Inf, sort.by="none"), Hday01 =topTable(fit2, coef=5, adjust="BH", number=Inf, sort.by="none"), Hday12 =topTable(fit2, coef=6, adjust="BH", number=Inf, sort.by="none"),  Hday23 =topTable(fit2, coef=7, adjust="BH", number=Inf, sort.by="none"),  Cday01 =topTable(fit2, coef=8, adjust="BH", number=Inf, sort.by="none"), Cday12 =topTable(fit2, coef=9, adjust="BH", number=Inf, sort.by="none"), Cday23 =topTable(fit2, coef=10, adjust="BH", number=Inf, sort.by="none"))

important_columns <- c(1,2,6)

# Find the genes that are DE at Day 0
HvCday0 =topTable(fit2, coef=1, adjust="BH", number=Inf, sort.by="none")
nrow(HvCday0[which(HvCday0$adj.P.Val < 0.05), important_columns])
HvCday0 <- HvCday0[which(HvCday0$adj.P.Val < 0.05), 1]             

# Find the genes that are DE at Day 1
HvCday1 =topTable(fit2, coef=2, adjust="BH", number=Inf, sort.by="none")
nrow(HvCday1[which(HvCday1$adj.P.Val < 0.05), important_columns])
HvCday1 <- HvCday1[which(HvCday1$adj.P.Val < 0.05), 1]

# Find the genes that are DE at Day 2
HvCday2 =topTable(fit2, coef=3, adjust="BH", number=Inf, sort.by="none")
nrow(HvCday2[which(HvCday2$adj.P.Val < 0.05), important_columns])
HvCday2 <- HvCday2[which(HvCday2$adj.P.Val < 0.05), 1]


# Find the genes that are DE at Day 3
HvCday3 =topTable(fit2, coef=4, adjust="BH", number=Inf, sort.by="none")
nrow(HvCday3[which(HvCday3$adj.P.Val < 0.05), important_columns])
HvCday3 <- HvCday3[which(HvCday3$adj.P.Val < 0.05), 1]


important_columns <- c(1,2,6)

# Find the genes that are DE at Human Day 0 to Day 1
H_day01 =topTable(fit2, coef=5, adjust="BH", number=Inf, sort.by="none")
dim(H_day01[which(H_day01$adj.P.Val < 0.05),])
H_day01 <- H_day01[, important_columns]    

# Find the genes that are DE at Human Day 1 to Day 2
H_day12 =topTable(fit2, coef=6, adjust="BH", number=Inf, sort.by="none")
H_day12 <- H_day12[, important_columns]  

# Find the genes that are DE at Human Day 2 to Day 3
H_day23 =topTable(fit2, coef=7, adjust="BH", number=Inf, sort.by="none")
H_day23 <- H_day23[, important_columns]  

# Find the genes that are DE at Chimp Day 0 to Day 1
C_day01 =topTable(fit2, coef=8, adjust="BH", number=Inf, sort.by="none")
C_day01 <- C_day01[, important_columns]    

# Find the genes that are DE at Chimp Day 1 to Day 2
C_day12 =topTable(fit2, coef=9, adjust="BH", number=Inf, sort.by="none")
C_day12 <- C_day12[, important_columns]  

# Find the genes that are DE at Chimp Day 2 to Day 3
C_day23 =topTable(fit2, coef=10, adjust="BH", number=Inf, sort.by="none")
C_day23 <- C_day23[, important_columns]  



# Check dimensions
dim(H_day01)
dim(H_day12)
dim(H_day23)
dim(C_day01)
dim(C_day12)
dim(C_day23)


mylist <- list()
mylist[["DE Day 0"]] <- HvCday0
mylist[["DE Day 1"]] <- HvCday1 
mylist[["DE Day 2"]] <- HvCday2
mylist[["DE Day 3"]] <- HvCday3


# Make as pdf 
Four_comp <- venn.diagram(mylist, filename= NULL, main="DE genes between species per day (5% FDR, RPKM, 63 samples)", cex=1.5 , fill = pal[1:4], lty=1, height=2000, width=3000)
pdf(file = "~/Dropbox/Endoderm TC/Tables_Supplement/Supplementary_Figures/SF4w_Four_comparisons_RPKM.pdf")
  grid.draw(Four_comp)
dev.off()


```


## Basic plot of length difference by DE status (RPKM values) 
```{r}
# Load data

#length_diff <- read.delim("~/Dropbox/Endoderm TC/length_diff_RPKM.txt")

# For each day

#p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot()

#p


# Each day plotted individually

#p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) + ylab("Human length -  chimp length (bp)") + xlab("DE based on 2 value cutoff")

#p

#q <- p + facet_wrap( ~ day) 
#q

# Zoom in

#p <- ggplot(length_diff, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) 

#q <- p + facet_wrap( ~ day) + ylim(c(-50,50))
#q

```

## Genes with difference > 10bp between species

```{r}

#length_diff_10 <- length_diff[which(length_diff$abld > 10) , ]

# Each day plotted individually

#p <- ggplot(length_diff_10, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=day)) + ylab("Human length -  chimp length (bp)") + xlab("DE (5% FDR)")

#q <- p + facet_wrap( ~ day) +geom_jitter(position=position_jitter(width=.1, height=0)) + ylim(c(-200,200))
#q


```

## Genes with difference > 10bp between species further divided into direction of effect size

```{r}
# Separate by direction

#length_diff_10$logFC[length_diff_10$logFC <= 0] <- -1
#length_diff_10$logFC[length_diff_10$logFC > 0] <- 1

#length_diff_10$inter <- interaction(length_diff_10$day, length_diff_10$logFC)

# Each day plotted individually

#p <- ggplot(length_diff_10, aes(factor(DEsig), lengthdiff)) + geom_boxplot(aes(fill=inter)) + ylab("Human length -  chimp length (bp)") + xlab("DE based on 2 value cutoff (neg. logFC / pos. logFC)")

#q <- p + facet_wrap( ~ day) +geom_jitter(position=position_jitter(width=0.8, height=1)) + ylim(c(-200,200))
#q
```


