---
title: "Endoderm_TC_Technical_Factor_Analysis"
author: "Lauren Blake"
date: "January 12, 2017"
output: html_document
---

The goal of this is to establish which, if any, technical factors are correlated with our biological variables of interest. 

# PART ONE: See if any of the variables for RNA-Seq correlate with the expression PCs for genes

## Initialization

```{r}
# Load libraries

library("polycor")
library("xlsx")

# Load cpm data 

cpm_in_cutoff <- read.delim("~/Desktop/Endoderm_TC/ashlar-trial/data/cpm_norm_data.txt")

# Load sample information

After_removal_sample_info <- read.csv("~/Desktop/Endoderm_TC/After_removal_sample_info.csv")

Species <- After_removal_sample_info$Species
species <- After_removal_sample_info$Species
day <- After_removal_sample_info$Day
individual <- After_removal_sample_info$Individual
Sample_ID <- After_removal_sample_info$Sample_ID
labels <- paste(Sample_ID, day, sep=" ")

# Load technical factor information



```

## Obtain gene expression PCs

```{r}
# Make PCA plots with the factors colored by day

pca_genes <- prcomp(t(mean_tech_reps), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
pc1 <- matrixpca[,1]
pc2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
