---
title: "Endoderm_TC_Technical_Factor_Analysis"
author: "Lauren Blake"
date: "January 12, 2017"
output: html_document
---

The goal of this is to establish which, if any, technical factors are correlated with our biological variables of interest. 

# PART ONE: See if any of the variables for RNA-Seq correlate with the expression PCs for genes (63 samples)

## Initialization

```{r}
# Load libraries

library("gdata")
library("ggplot2")
library("qvalue")
library("glmnet")
source("~/Desktop/Endoderm_TC/ashlar-trial/analysis/chunk-options.R")


# Load cpm data 

cpm_in_cutoff <- read.delim("~/Desktop/Endoderm_TC/ashlar-trial/data/cpm_cyclicloess.txt")

# Load sample information

After_removal_sample_info <- read.csv("~/Desktop/Endoderm_TC/After_removal_sample_info.csv")

Species <- After_removal_sample_info$Species
species <- After_removal_sample_info$Species
day <- After_removal_sample_info$Day
individual <- After_removal_sample_info$Individual
Sample_ID <- After_removal_sample_info$Sample_ID
labels <- paste(Sample_ID, day, sep=" ")

# Load technical factor information

RNA_seq_info_all <- read.csv("~/Desktop/Endoderm_TC/ashlar-trial/data/Endo_TC_Data_Share_Sorting.csv",  header = T)

dim(RNA_seq_info_all)

RNA_seq_info <- as.data.frame(cbind(RNA_seq_info_all[1:63, 4], RNA_seq_info_all[1:63, 3], RNA_seq_info_all[1:63, 5:27], RNA_seq_info_all[1:63, 30:35], RNA_seq_info_all[1:63, 37:43]))

# Remove day (biological variable of interest)
RNA_seq_info <- RNA_seq_info[,-8]

# Remove library well (only 1/well)
RNA_seq_info <- RNA_seq_info[,-19]

# Full data set
dim(RNA_seq_info)
```

## Obtain gene expression PCs

```{r}
# PCs

pca_genes <- prcomp(t(cpm_in_cutoff), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
pc1 <- matrixpca[,1]
pc2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]

pcs <- data.frame(pc1, pc2, pc3, pc4, pc5)

summary <- summary(pca_genes)

```

## Plots for PCs versus technical factors

```{r}
#Create plots for each of the possible confounders versus PCs 1-5

pdf('~/Desktop/Endoderm_TC/ashlar-trial/data/VarVsGenePCs.pdf')

for (i in 2:length(RNA_seq_info)) { 
  par(mfrow=c(1,5))
  plot(RNA_seq_info[,i], pcs[,1],  ylab = "PC1", xlab = " ")
  plot(RNA_seq_info[,i], pcs[,2],  ylab = "PC2", xlab = " ")
  plot(RNA_seq_info[,i], pcs[,3],  ylab = "PC3", xlab = " ")
  plot(RNA_seq_info[,i], pcs[,4],  ylab = "PC4", xlab = " ")
  plot(RNA_seq_info[,i], pcs[,5],  ylab = "PC5", xlab = " ")
  title(xlab = substitute(paste(k), list(k=colnames(RNA_seq_info)[i])), outer = TRUE, line = -2)
  mtext(substitute(paste('PCs vs. ', k), list(k=colnames(RNA_seq_info)[i])), side = 3, line = -2, outer = TRUE)

}
dev.off()
```

## Testing association between a particular variable and PCs with a linear model

```{r}
# TESTING BIOLOGICAL VARIABLES OF INTEREST

PC_pvalues_day = matrix(data = NA, nrow = 5, ncol = 1, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Day")))

for(i in 1:5){
# PC versus day

 checkPC1 <- lm(pcs[,i] ~ as.factor(day))

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  PC_pvalues_day[i,1] <- p_fstat
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared

}
# PC  versus species

PC_pvalues_species = matrix(data = NA, nrow = 5, ncol = 1, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Species")))

for(i in 1:5){
# PC versus species

   checkPC1 <- lm(pcs[,i] ~ as.factor(species))

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  PC_pvalues_species [i,1] <- p_fstat 
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared
}

# TESTING TECHNICAL VARIABLES OF INTEREST
  
#Make an empty matrix to put all of the data in
# Note: Do not include TC day, as it is a biological variable of interest

PC_pvalues = matrix(data = NA, nrow = 5, ncol = 35, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Cell line", "SPSX", "Batch", "Passage at seed",  "Start date", "Density at seed", "Harvest time", "Harvest density", "High conf purity", "Max purity", "RNA Extraction Date", "RNA conc", "RIN", "260 280 RNA", "260 230 RNA", "DNA concentration", "Library prep batch", "Library concentration", "uL sample", "uL EB", "Index sequence", "Seq pool", "Lane r1", "Mseqs R1", "Total lane reads 1", "Lane prop r1", "Dup r1", "GC r1", "Lane r2", "Mseqs r2", "Total lane reads r2", "Lane prop r2", "Dups r2", "GC r2", "Total reads")))

PC_r2 = matrix(data = NA, nrow = 5, ncol = 35, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Cell line", "SPSX", "Batch", "Passage at seed",  "Start date", "Density at seed",   "Harvest time", "Harvest density", "High conf purity", "Max purity", "RNA Extraction Date", "RNA conc", "RIN", "260 280 RNA", "260 230 RNA", "DNA concentration", "Library prep batch", "Library concentration", "uL sample", "uL EB", "Index sequence", "Seq pool", "Lane r1", "Mseqs R1", "Total lane reads 1", "Lane prop r1", "Dup r1", "GC r1", "Lane r2", "Mseqs r2", "Total lane reads r2", "Lane prop r2", "Dups r2", "GC r2", "Total reads")))



numerical_tech_factors <- c(5,7, 9:11, 13:17, 19:21, 25:29, 31:36)
categorical_tech_factors <- c(2:4,6, 8, 12, 18, 22:24, 30 )

#Check lm to see how well the variables containing numerical data are correlated with a PC

#For PCs 1-5
j=1
for (i in numerical_tech_factors){
  for (j in 1:5){
  
  checkPC1 <- lm(pcs[,j] ~ RNA_seq_info[,i])

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared

#Put the summary statistics into the matrix w

  PC_pvalues[j, i-1] <- p_fstat
  PC_r2[j, i-1] <- r2_value
  
  }

}
#Check lm to see how well the variables containing ordinal data are correlated with a PC

#For PCs 1-5
j=1
for (i in categorical_tech_factors){
  for (j in 1:5){
  
  checkPC1 <- lm(pcs[,j] ~ as.factor(RNA_seq_info[,i]))

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared

#Put the summary statistics into the matrix w

  PC_pvalues[j, i-1] <- p_fstat
  PC_r2[j, i-1] <- r2_value
  
  }

}

```

## Test for potential violations of the assumptions of the linear model

Note: I learned in http://lauren-blake.github.io/Reg_Evo_Primates/analysis/Tech_factor_analysis1_gene_exp.html that this doesn't work when one or more cells in a column contains an "NA". 
```{r}
#Plot the residuals to look for violations of the assumptions of lm

#pdf('~/Desktop/Endoderm_TC/ashlar-trial/data/Residuals_vs_var_GenePCs.pdf')


#non_NA_numerical_tech_factor <- c(5,8,  14, 20:22, 26:30, 32:37)
#non_NA_categorical_tech_factor <- c(2:4,6, 9, 13, 18:19, 23:25, 31 )

#for (i in non_NA_numerical_tech_factor){
  
#  par(mfrow=c(1,5))
  
#  checkPC1 <- lm(pcs[,1] ~ RNA_seq_info[,i])
#  plot(RNA_seq_info[,i], resid(checkPC1), ylab = "Residuals (lm with PC1)", xlab = " ")
  
#  checkPC2 <- lm(pcs[,2] ~ RNA_seq_info[,i])
#  plot(RNA_seq_info[,i], resid(checkPC2), ylab = "Residuals (lm with PC2)", xlab = " ")
  
#  checkPC3 <- lm(pcs[,3] ~ RNA_seq_info[,i])
#  plot(RNA_seq_info[,i], resid(checkPC3), ylab = "Residuals (lm with PC3)", xlab = #" ")
  
#  checkPC4 <- lm(pcs[,4] ~ RNA_seq_info[,i])
#  plot(RNA_seq_info[,i], resid(checkPC4), ylab = "Residuals (lm with PC4)", xlab = " ")
  
#  checkPC5 <- lm(pcs[,5] ~ RNA_seq_info[,i])
#  plot(RNA_seq_info[,i], resid(checkPC5), ylab = "Residuals (lm with PC5)", xlab = " ")
  
#  title(xlab = substitute(paste(k), list(k=colnames(RNA_seq_info)[i])), outer = TRUE, line = -2)
#  mtext(substitute(paste('Residuals vs. ', k), list(k=colnames(RNA_seq_info)[i])), side = 3, line = -2, outer = TRUE)
  
#}

#  for (i in non_NA_categorical_tech_factor){
#    par(mfrow=c(1,5))
   
#    checkPC1 <- lm(pcs[,1] ~ as.factor(RNA_seq_info[,i]))
#    plot(RNA_seq_info[,i], resid(checkPC1), ylab = "Residuals (lm with PC1)", xlab = " ")

#    checkPC2 <- lm(pcs[,2] ~ as.factor(RNA_seq_info[,i]))
#    plot(RNA_seq_info[,i], resid(checkPC2),  ylab = "Residuals (lm with PC2)", xlab = " ")
    
#    checkPC3 <- lm(pcs[,3] ~ as.factor(RNA_seq_info[,i]))
#    plot(RNA_seq_info[,i], resid(checkPC3),  ylab = "Residuals (lm with PC3)", xlab = " ")
    
#    checkPC4 <- lm(pcs[,4] ~ as.factor(RNA_seq_info[,i]))
#    plot(RNA_seq_info[,i], resid(checkPC4),  ylab = "Residuals (lm with PC4)", xlab = " ")
    
#    checkPC5 <- lm(pcs[,5] ~ as.factor(RNA_seq_info[,i]))
#    plot(RNA_seq_info[,i], resid(checkPC5),  ylab = "Residuals (lm with PC5)", xlab = " ")
    
#    title(xlab = substitute(paste(k), list(k=colnames(RNA_seq_info)[i])), outer = TRUE, line = -2)
#    mtext(substitute(paste('Residuals vs. ', k), list(k=colnames(RNA_seq_info)[i])), side = 3, line = -2, outer = TRUE)
    
#  }

#dev.off()
```

## Find which factors are statistically significant

```{r}
#Distribution of p-values adjusted by FDR not including species or tissue

fdr_val = p.adjust(PC_pvalues, method = "fdr", n = length(PC_pvalues))
fdr_val_order = fdr_val[order(fdr_val)]
hist(fdr_val_order, ylab = "BH-adjusted p-values", main = "Distribution of Benjamini and Hochberg adjusted p-values", breaks = 10)

# Number of values significant at 10% FDR 

fdr_val <- matrix(fdr_val, nrow = 5, ncol = 35)
matrix_fdr_val = matrix(data = fdr_val, nrow = 5, ncol = 35, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"),  c("Cell line", "SPSX", "Batch", "Passage at seed",  "Start date", "Density at seed", "Harvest time", "Harvest density", "High conf purity", "Max purity", "RNA Extraction Date", "RNA conc", "RIN", "260 280 RNA", "260 230 RNA", "DNA concentration", "Library prep batch", "Library concentration", "uL sample", "uL EB", "Index sequence", "Seq pool", "Lane r1", "Mseqs R1", "Total lane reads 1", "Lane prop r1", "Dup r1", "GC r1", "Lane r2", "Mseqs r2", "Total lane reads r2", "Lane prop r2", "Dups r2", "GC r2", "Total reads")))

# Number of values significant at 10% FDR 
sum(matrix_fdr_val <= 0.1)

#Get the coordinates of which variables/PC combinations are significant at FDR 10%

TorF_matrix_fdr <- matrix_fdr_val <=0.1
coor_to_check <- which(matrix_fdr_val <= 0.1, arr.ind=T)
coor_to_check <- as.data.frame(coor_to_check)

# Number of variables significant at 10% FDR  (note: off by 1 from column number in RNA_seq_info file; see notes in Part Two)

coor_to_check_col <- coor_to_check$col
unique(coor_to_check_col)
```

** Conclusions from Part I**

The following variables are associated with one of the PCs tested and will be investigated further in Part 2.

* Cell line
* SPSX
* Batch
* Passage at seed
* Start date 
* Harvest time
* Harvest density
* High conf purity
* Max purity
* RNA Extraction Date
* RNA conc
* RIN
* 260 280 RNA 
* Library prep batch 
* Index sequence
* Seq pool
* Lane r1
* Total lane reads 1
* Lane r2

# PART TWO: For the variable(s) that correlate, see if these segregate with either species or tissue

In coor_to_check_col, row is the PC # and col is the column # -1 that is associated with the PC. Want to take the coor_to_check_col column # and add one. 
```{r}
var_numb = unique(coor_to_check_col) + 1
var.numb <- as.data.frame(var_numb)

pdf('~/Desktop/Endoderm_TC/ashlar-trial/data/Var_GenePCsVsSpeciesTissue.pdf')

for (i in var.numb[1:nrow(var.numb),]) { 
par(mfrow=c(1,2))
plot(day, RNA_seq_info[,i], xlab = "Species", ylab = substitute(paste(k), list(k=colnames(RNA_seq_info)[i])))

plot(Species, RNA_seq_info[,i], xlab = "Tissue", ylab = substitute(paste(k), list(k=colnames(RNA_seq_info)[i])))

mtext(substitute(paste(k, ' vs. Species and Tissue'), list(k=colnames(RNA_seq_info)[i])), side = 3, line = -2, outer = TRUE)

}
dev.off()
```

## Testing to see if differences across variable groups are statistically significant for day.

```{r}
# Technical factors to be tested 

numerical_prev_sign_tech_factors <- c(5, 9:11, 13:15, 26)
categorical_prev_sign_tech_factors <- c(2:4,6, 8, 12, 18, 22:24, 30 )

#Make a matrix to store the p-values

pvalues_day = matrix(data = NA, nrow = 1, ncol = 35, dimnames = list(c("p-value"),  c("Cell line", "SPSX", "Batch", "Passage at seed",  "Start date", "Density at seed", "Harvest time", "Harvest density", "High conf purity", "Max purity", "RNA Extraction Date", "RNA conc", "RIN", "260 280 RNA", "260 230 RNA", "DNA concentration", "Library prep batch", "Library concentration", "uL sample", "uL EB", "Index sequence", "Seq pool", "Lane r1", "Mseqs R1", "Total lane reads 1", "Lane prop r1", "Dup r1", "GC r1", "Lane r2", "Mseqs r2", "Total lane reads r2", "Lane prop r2", "Dups r2", "GC r2", "Total reads")))

# Numerical

#Performing this test of significance for variables that are numerical data (Using an ANOVA)
j=1
for (i in numerical_prev_sign_tech_factors) {  
summary_anova = summary(aov(RNA_seq_info[,i]~ as.factor(day)))
data_summary_anova <- as.data.frame(summary_anova[[1]]$`Pr(>F)`)
p_val_anova <- data_summary_anova[1,]
pvalues_day[, i-1] <- p_val_anova  
j=j+1
}

# Ordinal

#Performing this test of significance for variables that are categorical data (Using Pearson's chi-squared test)
j=1
for (i in categorical_prev_sign_tech_factors) { 
  
  pval_chi <- chisq.test(as.factor(RNA_seq_info[,i]), as.factor(day), simulate.p.value = TRUE)$p.value
  pvalues_day[, i-1] <- pval_chi  
  j=j+1
  
}


```


## Testing to see if differences across variable groups are statistically significant for species.

```{r}
# Technical factors to be tested 

numerical_prev_sign_tech_factors <- c(5, 9:11, 13:15, 26)
categorical_prev_sign_tech_factors <- c(2:4,6, 8, 12, 18, 22:24, 30 )

#Make a matrix to store the p-values

pvalues_species = matrix(data = NA, nrow = 1, ncol = 35, dimnames = list(c("p-value"),  c("Cell line", "SPSX", "Batch", "Passage at seed",  "Start date", "Density at seed", "Harvest time", "Harvest density", "High conf purity", "Max purity", "RNA Extraction Date", "RNA conc", "RIN", "260 280 RNA", "260 230 RNA", "DNA concentration", "Library prep batch", "Library concentration", "uL sample", "uL EB", "Index sequence", "Seq pool", "Lane r1", "Mseqs R1", "Total lane reads 1", "Lane prop r1", "Dup r1", "GC r1", "Lane r2", "Mseqs r2", "Total lane reads r2", "Lane prop r2", "Dups r2", "GC r2", "Total reads")))

# Numerical

#Performing this test of significance for variables that are numerical data (Using an ANOVA. Note: in this case, the p-value of the ANOVA is the same as the p-value of the beta1 coefficient in lm)
j=1
for (i in numerical_prev_sign_tech_factors) {  
summary_anova = summary(aov(RNA_seq_info[,i]~ as.factor(Species)))
data_summary_anova <- as.data.frame(summary_anova[[1]]$`Pr(>F)`)
p_val_anova <- data_summary_anova[1,]
pvalues_species[, i-1] <- p_val_anova  
j=j+1
}


# Ordinal

#Performing this test of significance for variables that are categorical data (Using Pearson's chi-squared test)
j=1
for (i in categorical_prev_sign_tech_factors) { 
  
  pval_chi <- chisq.test(as.factor(RNA_seq_info[,i]), as.factor(Species), simulate.p.value = TRUE)$p.value
  pvalues_species[, i-1] <- pval_chi  
  j=j+1
  
}

# Combine tables

collapse_table_full <- rbind(pvalues_day, pvalues_species)

# Get rid of the NAs

collapse_table <- collapse_table_full[,(var_numb-1)]
collapse_table
```


## Find factors with FDR < 0.1

```{r}
#Calculate q-values (FDR = 10%)

fdr_val = p.adjust(collapse_table, method = "fdr", n = length(collapse_table)*2)
fdr_val_order = fdr_val[order(fdr_val)]
hist(fdr_val_order, ylab = "BH-adjusted p-values", main = "Distribution of Benjamini and Hochberg adjusted p-values", breaks = 10)

collapse_table_fdr_val = matrix(data = fdr_val, nrow = 2, ncol = nrow(var.numb), dimnames = list(c("Day", "Species"), colnames(collapse_table)))

collapse_table_fdr_val
```

** Conclusions from Part 2 **

The following variables are confounded with day:

* Harvest time (BH adjusted p-value = 0.009495252)
* Harvest density (BH adjusted p-value = 0.01327875)
* Index sequence (BH adjusted p-value = 0.009495252)

The following variables are confounded with species:

* Cell line (BH adjusted p-value = 0.009495252)
* SPSX (BH adjusted p-value = 0.009495252)
* Harvest density (BH adjusted p-value = 0.03300125)
* Max. purity (BH adjusted p-value = 0.05001969)

```{r}
plot(day, as.factor(RNA_seq_info[,2]), main = "Cell line versus day")
plot(Species, as.factor(RNA_seq_info[,2]), main = "Cell line versus species (sign.)")
plot(day, as.factor(RNA_seq_info[,3]), main = "SPSX versus day")
plot(Species, as.factor(RNA_seq_info[,3]), main = "SPSX versus species (sign.)")
plot(day, as.factor(RNA_seq_info[,8]), main =  "Harvest time versus day (sign.)")
plot(Species, RNA_seq_info[,8], main = "Harvest time versus species")

plot(as.factor(day), RNA_seq_info[,9], main = "Harvest density versus day (sign.)", ylab = "Harvest density", xlab = "Day")
plot(Species, RNA_seq_info[,9], main = "Harvest density versus species (sign.)", ylab = "Harvest density", xlab = "Species")

plot(as.factor(day), RNA_seq_info[,10], main = "High confidence purity versus day", ylab = "High confidence purity", xlab = "Day")
plot(Species, RNA_seq_info[,10], main = "High confidence purity versus species", ylab = "High confidence purity", xlab = "Species")
plot(as.factor(day), RNA_seq_info[,11], main = "Max. purity versus day",  ylab = "Max. purity", xlab = "Day")
plot(Species, RNA_seq_info[,11], main = "Max. purity versus species (sign.)", ylab = "Max. purity", xlab = "Species")
plot(as.factor(day), as.factor(RNA_seq_info[,22]), main = "Index sequence versus day (sign.)")
plot(Species, RNA_seq_info[,22], main = "Index sequence versus species")
```

### Combine and test day versus adapter for each sequencing pool

```{r}
# Make table of day, adaptor, and sequencing pool

day_index_pool <- cbind(day, RNA_seq_info[,22], RNA_seq_info[,23]) 

# Test dependency with the adaptors by day 

  # Sequencing pool 1
adaptors_pool1 <- day_index_pool[ which(day_index_pool[,3]==1),]

chisq.test(as.factor(adaptors_pool1[,2]), as.factor(adaptors_pool1[,1]), simulate.p.value = TRUE)$p.value

#fisher.test(adaptors_pool1[,1:2], simulate.p.value=TRUE)$p.value
#fisher.test(adaptors_pool2[,2:1], simulate.p.value=TRUE)$p.value
#fisher.test(adaptors_pool3[,2:1], simulate.p.value=TRUE)$p.value
#fisher.test(adaptors_pool4[,2:1], simulate.p.value=TRUE)$p.value

  # p-value = 1

  # Sequencing pool 2
adaptors_pool2 <- day_index_pool[ which(day_index_pool[,3]==2),]

chisq.test(as.factor(adaptors_pool2[,2]), as.factor(adaptors_pool2[,1]), simulate.p.value = TRUE)$p.value

  # p-value = 1

  # Sequencing pool 3
adaptors_pool3 <- day_index_pool[ which(day_index_pool[,3]==3),]

chisq.test(as.factor(adaptors_pool3[,2]), as.factor(adaptors_pool3[,1]), simulate.p.value = TRUE)$p.value

  # p-value = 1

  # Sequencing pool 4
adaptors_pool4 <- day_index_pool[ which(day_index_pool[,3]==4),]

chisq.test(as.factor(adaptors_pool4[,2]), as.factor(adaptors_pool4[,1]), simulate.p.value = TRUE)$p.value

  # p-value = 1

fdr_val_4= p.adjust(c(1,1,1,1), method = "fdr")
```




## PART THREE: Which variables to put in the model?

### Check correlation between harvest density and max. purity

Check for degree of similarity between the values of the different technical variables, harvest density and max. purity. For ones that show high correlation (multi-colinearity), we will only want to incorporate one of these variables into the model. 

```{r}
# Correlation bet. harvest density and max. purity.
  corr_tech <- summary(lm(RNA_seq_info[,9] ~ RNA_seq_info[,11]))


#Get the p-value of the F-statistic
  corr_tech$fstatistic

  fstat <- as.data.frame(corr_tech$fstatistic)
  1-pf(fstat[1,], fstat[2,], fstat[3,])
  

```

These two factors are not highly correlated (p-value = 0.465). 

### RIN Score 

```{r}
# Make a table with species, day, and RIN score
species <- After_removal_sample_info$Species
day <- After_removal_sample_info$Day

iPSC_prop <- as.data.frame(cbind(day, species, RNA_seq_info[,14]), stringsAsFactors = FALSE)

# Remove the 1 sample with a missing RIN score

remove_NA <- c(26)

iPSC_prop <- iPSC_prop[-remove_NA, ]

colnames(iPSC_prop) <- c("Day", "Species", "RIN_Score") 

iPSC_prop$Species[iPSC_prop$Species == "2"] <- "Human"
iPSC_prop$Species[iPSC_prop$Species == "1"] <- "Chimp"

iPSC_prop$Day[iPSC_prop$Day == "0"] <- "Day 0"
iPSC_prop$Day[iPSC_prop$Day == "1"] <- "Day 1"
iPSC_prop$Day[iPSC_prop$Day == "2"] <- "Day 2"
iPSC_prop$Day[iPSC_prop$Day == "3"] <- "Day 3"

ggplot(data = iPSC_prop, aes(y = RIN_Score, x = as.factor(Species)))  + facet_wrap(~ Day, nrow = 1) + geom_boxplot() + geom_point(size = 5,  position=position_jitter(width=0.2, height=0.1), show.legend = FALSE) + labs(x = "Species", y = "RIN score") + theme_bw()


ggplot(data = iPSC_prop, aes(y = RIN_Score, x = as.factor(Species)))  + facet_wrap(~ Day, nrow = 1) + geom_boxplot()  + geom_dotplot(binaxis='y', stackdir='center', binwidth = 0.05) + labs(x = "Species", y = "RIN score") + theme_bw()


```



### Harvest time and harvest densities (supplement)

```{r}
iPSC_prop <- as.data.frame(cbind(day, species, RNA_seq_info[,9],  as.character(RNA_seq_info[,8])), stringsAsFactors = FALSE)

  
colnames(iPSC_prop) <- c("Day", "Species", "Harvest_density", "Harvest_time") 

iPSC_prop$Species[iPSC_prop$Species == "2"] <- "Human"
iPSC_prop$Species[iPSC_prop$Species == "1"] <- "Chimp"

iPSC_prop$Day[iPSC_prop$Day == "0"] <- "Day 0"
iPSC_prop$Day[iPSC_prop$Day == "1"] <- "Day 1"
iPSC_prop$Day[iPSC_prop$Day == "2"] <- "Day 2"
iPSC_prop$Day[iPSC_prop$Day == "3"] <- "Day 3"

# Time is 7

remove_NA <- c(7,23,39,55)

iPSC_prop <- iPSC_prop[-remove_NA, ]

iPSC_prop$Harvest_density <- as.numeric(iPSC_prop$Harvest_density)

ggplot(data = iPSC_prop, aes(y = Harvest_density, x = as.factor(Day)))  + facet_wrap(~ Species, nrow = 1) + geom_boxplot() + geom_point(aes(color = as.factor(Species)), size = 5,  position=position_jitter(width=0.2, height=0.1), show.legend = FALSE)

ggplot(data = iPSC_prop, aes(y = Harvest_density, x = as.factor(Species)))  + facet_wrap(~ Day, nrow = 1) + geom_boxplot() 

# Harvest density (supplement)

ggplot(data = iPSC_prop, aes(y = Harvest_density, x = Species))  + facet_wrap(~ Day, nrow = 1) + geom_boxplot() + geom_dotplot(binaxis='y', stackdir='center') + labs(y = "Harvest density", title = "Harvest densities for each sample") + theme_bw()

ggplot(data = iPSC_prop, aes(y = Harvest_density, x = Species))  + facet_wrap(~ Day, nrow = 1) + geom_boxplot() + geom_point(size = 5,  position=position_jitter(width=0.2, height=0.1), show.legend = FALSE) + labs(y = "Harvest density", title = "Harvest densities for each sample") + theme_bw()


# Harvest density for each day species

median(iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "human") , 3])
median(iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "chimp") , 3])

median(iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "human") , 3])
median(iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "chimp") , 3])

median(iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "human") , 3])
median(iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "chimp") , 3])

median(iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "human") , 3])
median(iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "chimp") , 3])

# Harvest time

species <- After_removal_sample_info$Species
day <- After_removal_sample_info$Day

iPSC_prop <- as.data.frame(cbind(day, species, RNA_seq_info[,9],  as.character(RNA_seq_info[,8])), stringsAsFactors = FALSE)

colnames(iPSC_prop) <- c("Day", "Species", "Harvest_density", "Harvest_time") 

iPSC_prop$Species[iPSC_prop$Species == "2"] <- "Human"
iPSC_prop$Species[iPSC_prop$Species == "1"] <- "Chimp"

iPSC_prop$Day[iPSC_prop$Day == "0"] <- "Day 0"
iPSC_prop$Day[iPSC_prop$Day == "1"] <- "Day 1"
iPSC_prop$Day[iPSC_prop$Day == "2"] <- "Day 2"
iPSC_prop$Day[iPSC_prop$Day == "3"] <- "Day 3"

table(iPSC_prop$Day, iPSC_prop$Harvest_time)

ggplot(data = iPSC_prop, aes(y = as.factor(Harvest_time), x = Species))  + facet_wrap(~ Day, nrow = 1)  + geom_point(aes(color = Species), size = 5,  position=position_jitter(width=0.2, height=0.1), show.legend = FALSE) + theme_bw()

iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "chimp") , 4]
d0c <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "chimp") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "human") , 4]
d0h <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 0" & iPSC_prop$Species == "human") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "chimp") , 4]
d1c <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "chimp") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "human") , 4]
d1h <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 1" & iPSC_prop$Species == "human") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "chimp") , 4]
d2c <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "chimp") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "human") , 4]
d2h <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 2" & iPSC_prop$Species == "human") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "chimp") , 4]
d3c <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "chimp") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "human") , 4]
d3h <- table(factor(iPSC_prop[which(iPSC_prop$Day == "Day 3" & iPSC_prop$Species == "human") , 4], levels = c("11:56", "12:10", "12:18", "12:30", "1:00 PM", "1:05 PM", "1:06 PM", "1:08 PM", "1:17 PM", "1:20 PM", "1:30 PM")))

#Check that the frequencies match the input (n = 59)

frequency_time <- as.data.frame(cbind(d0c, d0h, d1c, d1h, d2c, d2h, d3c, d3h))

frequency_time

plot(frequency_time[,1])

rowSums(frequency_time)

# Order correctly

iPSC_prop$Harvest_time[iPSC_prop$Harvest_time == "11:56"] <- "11:56 AM"
iPSC_prop$Harvest_time[iPSC_prop$Harvest_time == "12:10"] <- "12:10 PM"
iPSC_prop$Harvest_time[iPSC_prop$Harvest_time == "12:18"] <- "12:18 PM"
iPSC_prop$Harvest_time[iPSC_prop$Harvest_time == "12:30"] <- "12:30 PM"

lv <- c("1:30 PM", "1:20 PM", "1:17 PM", "1:08 PM", "1:06 PM", "1:05 PM", "1:00 PM", "12:30 PM", "12:18 PM", "12:10 PM", "11:56 AM")
  
  
x <- factor(iPSC_prop$Harvest_time,levels = lv)

# Make plot

ggplot(data = iPSC_prop, aes(y = as.factor(x), x = as.factor(Species)))  + facet_wrap(~ Day, nrow = 1)  + geom_dotplot(binaxis='y', stackdir='center', binwidth = 0.2) + labs(x = "Species", y = "Harvest time") + theme_bw()

```



# PART ONE: See if any of the variables for RNA-Seq correlate with the expression PCs for genes (40 samples)

## Initialization

```{r}
# Load cpm data 

cpm_in_cutoff_40 <- read.delim("~/Desktop/Endoderm_TC/ashlar-trial/data/cpm_cyclicloess_40.txt")

# Load sample information

bio_rep_samplefactors <- read.delim("~/Desktop/Endoderm_TC/ashlar-trial/data/samplefactors-filtered.txt", stringsAsFactors=FALSE)
day <- bio_rep_samplefactors$Day
species <- bio_rep_samplefactors$Species
```

## Obtain gene expression PCs (40 samples)

```{r}
# PCs

pca_genes <- prcomp(t(cpm_in_cutoff_40), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
pc1 <- matrixpca[,1]
pc2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]

pcs <- data.frame(pc1, pc2, pc3, pc4, pc5)

summary <- summary(pca_genes)

```


## Testing association between a particular variable and PCs with a linear model (40 samples)

```{r}
# TESTING BIOLOGICAL VARIABLES OF INTEREST

PC_pvalues_day = matrix(data = NA, nrow = 5, ncol = 1, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Day")))

for(i in 1:5){
# PC versus day

 checkPC1 <- lm(pcs[,i] ~ as.factor(day))

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  PC_pvalues_day[i,1] <- p_fstat
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared

}
# PC  versus species

PC_pvalues_species = matrix(data = NA, nrow = 5, ncol = 1, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Species")))

for(i in 1:5){
# PC versus species

   checkPC1 <- lm(pcs[,i] ~ as.factor(species))

#Get the summary statistics from it
  summary(checkPC1)

#Get the p-value of the F-statistic
  summary(checkPC1)$fstatistic

  fstat <- as.data.frame(summary(checkPC1)$fstatistic)
  p_fstat <- 1-pf(fstat[1,], fstat[2,], fstat[3,])
  PC_pvalues_species [i,1] <- p_fstat 
  
#Fraction of the variance explained by the model
  r2_value <- summary(checkPC1)$r.squared
}

 # Combine tables

collapse_table_full <- rbind(PC_pvalues_day, PC_pvalues_species)

#Calculate q-values (FDR = 10%)

fdr_val = p.adjust(collapse_table_full, method = "fdr", n = length(collapse_table_full)*2)

collapse_table_fdr_val = matrix(data = fdr_val, nrow = 5, ncol = 2, dimnames = list(c("PC1", "PC2", "PC3", "PC4", "PC5"), c("Day", "Species")))

collapse_table_fdr_val
```

